Semantic Prediction in {R}ussian

### 2020-04-11 paper.tex
drop TODO; its task disappeared
fix missing enumerate, beautify items
fix up labels and tables

\documentclass[a4paper]{article}
\usepackage{lrec}
\usepackage{amsmath}
\usepackage{polyglossia}
\usepackage{graphicx}
\usepackage{hyperref}

@@

\begin{table*}
\caption{Example of stimuli sentence from RNC in cloze task with probabilities of the next word. Stimuli: \textit{"А промывать манную крупу перед тем, как варить ее, не пробовали?"} (English translation: \textit{"Have you tried to rinse semolina before boiling it?"})
\newline}
\centering
\begin{tabular}{llr}

@@

We used the dataset with cloze task answers from \cite{laurinavichyute}. The dataset is based on 144 sentences randomly selected from the National Corpus of the Russian Language (RNC, ruscorpora.ru) - an online corpus of Russian texts with extensive search options. These sentences were slightly edited: the authors replaced rare infrequent words with more frequent ones and shortened the sentences when they exceeded the preset maximum length of 13 words. The stimuli sentences were subjected to the cloze task experiment. Respondents were asked to successively predict the next words for each context. An example of stimuli that were shown to the participants is presented in Table 1 with corresponding correct next words and calculated predictability scores.

@@

Some corpus statistics are presented in Table 2.

\begin{table}
\caption{Corpus statistics. RNC is the Russian News Corpus, and NCRL is the National Corpus of the Russian Language.
\newline}
\centering

@@

\begin{enumerate}[1)]

\item        Extracts first 10 words (we have decided that that is the maximum amount of words for each context that would have meaningful probabilities as all of the words after the first ten for each model have their probabilities tends to zero) for each of 1219 contexts;

% TODO arithmetic should be TeX-ified
\item        Computes the mean probability of remaining words;

\item        Counts the difference between the probabilities of the first and last word and then the difference between mean probabilities of the previous word and the next one;

\item        Decides what amount of words vectors to use for each context based on how different the first-last word difference and the mean probability difference is – if the latter is lesser than the former, the function checks the next difference, if not – the previous amount of words would be used for semantic comparison.

\end{enumerate}

@@

\begin{tabular}{l}
\begin{russian} 1. Какие главные лекарства должны входить [в]\end{russian} \\
\begin{english} (What are the main drugs that should be included [in])\end{english} \\
\begin{russian} 2. В современном обществе семья и школа оказывают большое [влияние]\end{russian} \\
\begin{english} (In modern society, family and school have a large [influence])\end{english} \\
\begin{russian} 3. Зачем ему звонить если откликается спокойный женский [голос]\end{russian} \\
\begin{english} (Why would he call if a calm female [voice] answers)\end{english} \\
\begin{russian} 4. Ирине досталась отдельная комната в двухкомнатной [квартире]\end{russian} \\
\begin{english} (Irina got a separate room in a two-room [apartment])\end{english} \\
\begin{russian} 5. Они не ели целый [день]\end{russian} \\
\begin{english} (They haven't eaten all [day])\end{english} \\
\begin{russian} 6. Во избежание ожогов надо нанести на лицо небольшое [количество]\end{russian} \\
\begin{english} (To avoid burns on the face, apply a small [amount])\end{english} \\
\begin{russian} 7. Дрозды и скворцы начали вить семейные гнезда неподалеку друг от [друга]\end{russian} \\
\begin{english} (Blackbirds and starlings began to twist family nests not far from each [other])\end{english} \\
\begin{russian} 8. Собаку виновницу случившегося приказали сечь хотя в чем была ее [вина]\end{russian} \\
\begin{english} (The dog responsible for the incident was ordered to be beaten, although it wasn't really her [fault])\end{english} \\
\begin{russian} 9. С нескрываемой едкой иронией отзываются они друг о [друге]\end{russian} \\
\begin{english} (With undisguised caustic irony, they speak of each [other])\end{english} \\
\begin{russian} 10. Олень бродил среди берез жевал талый [снег]\end{russian} \\
\begin{english} (Deer wandered among the birches chewing melting [snow])\end{english} \\
\end{tabular}
\end{table*}

In Table 3, contexts with minimum variance in filling
are listed. It is worth emphasizing that in all these cases there was no variance in respondents’ answers, i.e. all respondents gave the same one answer for these contexts. What is more, this predicted word was the original word from the corpus.

@@

At the next step, we performed Kolmogorov-Smirnov testing to find the similarity in the probability distributions of the model predictions. Figure 5 also reflects the sum value of the pairwise comparison of the contexts using Kolmogorov-Smirnov testing. For convenience, all the values were normalized. Observations show that the closest probability distribution is seen in the LSTM model. However, as we see, there is a variation in metric values from 0.9 to 6.2 for different models. Thus, we can assert that the difference between all models and a cloze task is rather large and in some way unacceptable.

@@

To measure model predictions on the semantic level, the cosine similarity between each context’s predicted words centroid vector was found. The number of words was selected dynamically for each context by maximizing vector significance with the minimum words. Figure 6 reflects the results.
\begin{figure}
\caption{Cosine Similarity between each pair of models .}
\label{fig:ks-lexical-pairs}
\centering
\includegraphics[width=0.5\textwidth]{figures/pdf/cosine-similarity-lexical.pdf}
\end{figure}

@@

Table 3 shows the scores for three different models with
bigram HMM trained on NCRL showing the best results. Unfortunately, this heat map does not show us changes in the language model (LM) distances context-lengthwise. Top 3 lengthwise LM distances are showb in Table 4.

\begin{table}
\centering

@@

As we see, the bigram models start to higher the distance from 1 to 6 context length, but BERT, on the contrary, starts to lower the distance up to 6 context length. For this case, we merged the bigram model and BERT at length equal to 6.

KL(Cloze || Cloze) = 0.0

KL(Cloze || HMM) = 1.79

KL(Cloze || BERT) = 1.93

KL(Cloze || LSTM) = 1.84

KL(Cloze || HMM + BERT) = 1.71

@@

Consistency at the object-verb-functional-modifier level  and Kolmogorov-Smirnov Test are is shown in Figure 12 and 13 correspondingly.

>>>>

\documentclass[a4paper]{article}
\usepackage{lrec}
\usepackage{amsmath}
\usepackage{polyglossia}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}

@@

\begin{table*}
\label{tbl:stimuli}
\caption{Example of stimuli sentence from RNC in cloze task with
probabilities of the next word. Stimuli: \textit{``А промывать манную
крупу перед тем, как варить ее, не пробовали?''} (English translation:
\textit{``Have you tried to rinse semolina before boiling it?''})
\newline}
\centering
\begin{tabular}{llr}

@@

We used the dataset with cloze task answers from \cite{laurinavichyute}. The dataset is based on 144 sentences randomly selected from the National Corpus of the Russian Language (RNC, ruscorpora.ru) - an online corpus of Russian texts with extensive search options. These sentences were slightly edited: the authors replaced rare infrequent words with more frequent ones and shortened the sentences when they exceeded the preset maximum length of 13 words. The stimuli sentences were subjected to the cloze task experiment. Respondents were asked to successively predict the next words for each context. An example of stimuli that were shown to the participants is presented in Table~\ref{tbl:stimuli} with corresponding correct next words and calculated predictability scores.

@@

Some corpus statistics are presented in Table~\ref{tbl:corpus}.

\begin{table}
\label{tbl:corpus}
\caption{Corpus statistics. RNC is the Russian News Corpus, and NCRL is the National Corpus of the Russian Language.
\newline}
\centering

@@

\begin{enumerate}[1)]
\item Extracts first 10 words (we have decided that that is the maximum amount of words for each context that would have meaningful probabilities as all of the words after the first ten for each model have their probabilities tends to zero) for each of 1219 contexts;
\item Computes the mean probability of remaining words; 
\item Counts the difference between the probabilities of the first and last word and then the difference between mean probabilities of the previous word and the next one;
\item Decides what amount of words vectors to use for each context based on how different the first-last word difference and the mean probability difference is – if the latter is lesser than the former, the function checks the next difference, if not – the previous amount of words would be used for semantic comparison.
\end{enumerate}

@@

\begin{tabular}{rl}
1. & \begin{russian} Какие главные лекарства должны входить (в)\end{russian} \\
& \textit{ What are the main drugs that should be included (in) } \\
2. & \begin{russian} В современном обществе семья и школа оказывают большое (влияние)\end{russian} \\
& \textit{ In modern society, family and school have a large (influence) } \\
3. & \begin{russian} Зачем ему звонить если откликается спокойный женский (голос)\end{russian} \\
& \textit{ Why would he call if a calm female (voice) answers } \\
4. & \begin{russian} Ирине досталась отдельная комната в двухкомнатной (квартире)\end{russian} \\
& \textit{ Irina got a separate room in a two-room (apartment) } \\
5. & \begin{russian} Они не ели целый (день)\end{russian} \\
& \textit{ They haven't eaten all (day) } \\
6. & \begin{russian} Во избежание ожогов надо нанести на лицо небольшое (количество)\end{russian} \\
& \textit{ To avoid burns on the face, apply a small (amount) } \\
7. & \begin{russian} Дрозды и скворцы начали вить семейные гнезда неподалеку друг от (друга)\end{russian} \\
& \textit{ Blackbirds and starlings began to twist family nests not far from each (other) } \\
8. & \begin{russian} Собаку виновницу случившегося приказали сечь хотя в чем была ее (вина)\end{russian} \\
& \textit{ The dog responsible for the incident was ordered to be beaten, although it wasn't really her (fault) } \\
9. & \begin{russian} С нескрываемой едкой иронией отзываются они друг о (друге)\end{russian} \\
& \textit{ With undisguised caustic irony, they speak of each (other) } \\
10. & \begin{russian} Олень бродил среди берез жевал талый (снег)\end{russian} \\
& \textit{ The deer wandered among the birches chewing melting (snow) } \\
\end{tabular}
\end{table*}

In Table~\ref{tbl:min-var}, contexts with minimum variance in filling are listed. It is worth emphasizing that in all these cases there was no variance in respondents’ answers, i.e. all respondents gave the same one answer for these contexts. What is more, this predicted word was the original word from the corpus.

@@

At the next step, we performed Kolmogorov-Smirnov testing to find the similarity in the probability distributions of the model predictions. Figure~\ref{fig:ks-lexical} also reflects the sum value of the pairwise comparison of the contexts using Kolmogorov-Smirnov testing. For convenience, all the values were normalized. Observations show that the closest probability distribution is seen in the LSTM model. However, as we see, there is a variation in metric values from 0.9 to 6.2 for different models. Thus, we can assert that the difference between all models and a cloze task is rather large and in some way unacceptable.

@@

To measure model predictions on the semantic level, the cosine similarity between each context’s predicted words centroid vector was found. The number of words was selected dynamically for each context by maximizing vector significance with the minimum words. Figure~\ref{fig:cosine-lexical} reflects the results.
\begin{figure}
\caption{Cosine Similarity between each pair of models .}
\label{fig:cosine-lexical}
\centering
\includegraphics[width=0.5\textwidth]{figures/pdf/cosine-similarity-lexical.pdf}
\end{figure}

@@

Table~\ref{tbl:kl-scores} shows the scores for three different models with
bigram HMM trained on NCRL showing the best results. Unfortunately, this heat map does not show us changes in the language model (LM) distances context-lengthwise. Top 3 lengthwise LM distances from cloze are shown in Table~\ref{tbl:lm-distances}.

\begin{table}
\label{tbl:kl-scores}
\centering

@@

As we see, the bigram models start to increase the distance from 1 to 6 context length, but BERT, on the contrary, starts to lower the distance up to 6 context length. For this case, we merged the bigram model and BERT at length equal to 6.

\begin{table}
\label{tbl:lm-distances}
\centering

\caption{Distance in the KL-metric between the cloze task and language models.
\newline}

\begin{tabular}{lr}
\textbf{Model} &
\textbf{KL-distance to cloze} \\
\hline
   HMM & 1.79
\\ BERT & 1.93
\\ LSTM & 1.84
\\ HMM + BERT & 1.71
\end{tabular}
\end{table}

@@

Consistency at the object-verb-functional-modifier level  and Kolmogorov-Smirnov Test are is shown in Figure~\ref{fig:heatmap-context-accord-ovfm} and~\ref{fig:heatmap-ks-ovfm} correspondingly.
